# Summarizing Chess Tournament Info using R and Regular Expressions
### Heather Geiger
### February 23, 2018

## Reading in the data

To start, we read in the data in a way that each line is treated as one string.

Then, we use as.vector to make the result a vector instead of a data frame.

```{r}
#Set working directory to the directory you are currently in, which presumably will be where you downloaded this script and the text file.
setwd(getwd())
#Read in text file.
info <- as.vector(read.table("tournamentinfo.txt",quote="",comment.char="",sep="\n")$V1)
length(info)
head(info)
tail(info)
```

## Light pre-processing

Now, remove the lines that are all dashes.

For this we use the grep function, returning content of lines NOT matching a regex for a line of all dashes.

```{r}
library(stringr)
info <- grep('^-+$',info,invert=TRUE,perl=TRUE,value=TRUE)
length(info)
```

Set aside the first two items in info, as these were the headers.

```{r}
info_headers <- info[1:2]
info <- info[3:length(info)]
```

Really every two lines were a record. Let's separate the first line in each record from the second line.

```{r}
first_lines_of_records <- info[seq(from=1,to=(length(info) - 1),by=2)]
second_lines_of_records <- info[seq(from=2,to=length(info),by=2)]
length(first_lines_of_records);head(first_lines_of_records);tail(first_lines_of_records)
length(second_lines_of_records);head(second_lines_of_records);tail(second_lines_of_records)
```

## Separating by field and selecting the correct fields

Now, we want to get each field from first_lines_of_records and second_lines_of_records.

Then, the information we want will be contained within different fields in either first line or second line.

Fields are separated by dashes.

```{r}
first_lines_of_records_separate_fields <- str_split(first_lines_of_records,"\\|")
second_lines_of_records_separate_fields <- str_split(second_lines_of_records,"\\|")

#The first field from first lines of records gives the player number.
#Remove leading and trailing whitespace using base R trimws function to get the number by itself.
#In this case player_ids are in numerical order, but this might not always be the case so good to do.

player_ids <- unlist(lapply(first_lines_of_records_separate_fields,"[[",1))
player_ids <- trimws(player_ids)

#Repeat similar procedure for other fields.
#For some others like name, we'll need to do some additional processing after. 
#But at least good to extract each field into a vector for now.

player_names <- unlist(lapply(first_lines_of_records_separate_fields,"[[",2))
player_names <- trimws(player_names)

player_states <- trimws(unlist(lapply(second_lines_of_records_separate_fields,"[[",1)))

total_points <- trimws(unlist(lapply(first_lines_of_records_separate_fields,"[[",3)))

ratings <- trimws(unlist(lapply(second_lines_of_records_separate_fields,"[[",2)))

#Show all these vectors.

player_ids
player_names
player_states
total_points
ratings
```

## Processing the name field

The name and ratings fields require additional processing.

Let's focus on the name field first.

We'll want to have only the first letter of names be capitalized instead of the whole name.

It might be nice to add a period after initials and suffixes like Jr.

Finally, it seems there is a space in between for some last names that have two capitalized letters (like MC CORMICK for McCormick).

There also seems to be an error for one name. Assuming MC CORMICK is actually for McCormick, MCCLURE should be MC CLURE.

Let's start with last names.

```{r}
#First step - fix the McClure error.

player_names <- str_replace(player_names,pattern='MCCLURE',replace='MC CLURE')

#Now, create a version of player_names split by whitespace.
player_names_split <- str_split(player_names,"[[:blank:]]+")

#Now, we are going to create a few logical vectors looking for various things that will affect how we paste together names.
#Start by checking the number of "words" in each name. A lot of the tests we need to run only apply if there are more than two.

number_name_fields <- unlist(lapply(player_names_split,function(x)length(x)))

#If there are 3+ names, check the next-to-last for "MC".
#Extract next-to-last name using the "word" function.

two_capitals_last_name <- rep(FALSE,times=length(player_names))
two_capitals_last_name[which(number_name_fields >= 3 & word(player_names,-2) == "MC")] <- TRUE

#Next, check for Jr. suffix, as well as hypenated last name.
junior_suffix <- rep(FALSE,times=length(player_names))
junior_suffix[which(number_name_fields >= 3 & word(player_names,-1) == "JR")] <- TRUE

hyphenated_last_name <- str_detect(word(player_names,-1),'\\-')

#Now, let's get each last name with only the appropriate letters capitalized.
#For Jr. suffix, paste this to the second-to-last "word" in the name.
#Use the capitalize function from the Hmisc package to capitalize the first letter of each "word".
#Capitalize function used as suggested here: https://stackoverflow.com/questions/6364783/capitalize-the-first-letter-of-both-words-in-a-two-word-string.
#Combining this with the "tolower" function so only the first letter capitalized instead of the whole thing.

library(Hmisc)

last_names <- capitalize(tolower(word(player_names,-1)))
last_names[which(two_capitals_last_name == TRUE)] <- paste0("Mc",last_names[which(two_capitals_last_name == TRUE)])
last_names[which(junior_suffix == TRUE)] <- paste0(capitalize(tolower(word(player_names[which(junior_suffix == TRUE)],-2)))," Jr.")
for(i in which(hyphenated_last_name == TRUE))
{
this_last_name_split_by_hyphen <- unlist(str_split(last_names[i],"-"))
last_names[i] <- paste0(this_last_name_split_by_hyphen[1],"-",capitalize(this_last_name_split_by_hyphen[2]))
}

#Let's check our work.

data.frame(Full.name = player_names,Last.name = last_names)

```

Looks good! Now we need to get just the first and middle names/initials.

```{r}
first_names <- capitalize(tolower(word(player_names,1)))

#Adjust number_name_fields to subtract one for two_capitals_last_name or junior_suffix.
number_name_fields_adjusted <- number_name_fields
number_name_fields_adjusted[which(two_capitals_last_name == TRUE | junior_suffix == TRUE)] <- number_name_fields_adjusted[which(two_capitals_last_name == TRUE | junior_suffix == TRUE)] - 1

#After adjustment, how many name fields total are there?

table(number_name_fields_adjusted)

#Looks like no one has two middle names.
#So we can either say that there is no middle name or initial if number_name_fields_adjusted = 2, or there is a middle name or initial if number_name_fields_adjusted = 3.

#If number_name_fields_adjusted = 3, the middle name or initial will be the second name.

middle_names_or_initials <- rep("",times=length(player_names))
middle_names_or_initials[which(number_name_fields_adjusted == 3)] <- word(player_names[which(number_name_fields_adjusted == 3)],2)

#If middle is just a single letter, assume it is an initial and add a period.
#Otherwise, capitalize only the first letter.

middle_names_or_initials[which(nchar(middle_names_or_initials) == 1)] <- paste0(middle_names_or_initials[which(nchar(middle_names_or_initials) == 1)],".")
middle_names_or_initials[which(nchar(middle_names_or_initials) > 1)] <- capitalize(tolower(middle_names_or_initials[which(nchar(middle_names_or_initials) > 1)]))

middle_names_or_initials
```

Think we are good to go! Combine the first/middle/last information.

Then check everything at once versus the original names to see if it all looks right.

```{r}
player_names_reformatted <- paste0(first_names," ",middle_names_or_initials," ",last_names)
player_names_reformatted

#Looks like an extra space gets added if there is no middle name or initial. Let's fix this.

player_names_reformatted[which(nchar(middle_names_or_initials) == 0)] <- paste0(first_names[which(nchar(middle_names_or_initials) == 0)]," ",last_names[which(nchar(middle_names_or_initials) == 0)])

data.frame(Original = player_names,Reformat = player_names_reformatted)
```

## Processing the ratings field

For the ratings field, we can simplify a lot by taking everything before the arrow "->".

This is because we are interested in pre-ratings only.

Once we get everything before the "->", take the last part of THAT, which will always be after a colon.

```{r}
ratings <- str_split(ratings,'->')
ratings <- unlist(lapply(ratings,"[[",1))
ratings <- trimws(ratings,which="right")
ratings
ratings <- str_split(ratings,":")
ratings <- trimws(unlist(lapply(ratings,"[[",2)))
ratings
```

Finally, if there is any provisional notation, remove this by taking only the part before the "P" where appropriate.

```{r}
ratings <- str_split(ratings,"P")
ratings <- unlist(lapply(ratings,"[[",1))
ratings
```

Looks good! We'll need to convert to numeric, but we can do that later when we are assembling the data frame.

## Getting opponent player numbers

We'll need to process the remaining fields in first_lines_of_records_separate_fields (4th and onward) to get the player numbers of each player's opponents.

First, get just these values from first_lines_of_records_separate_fields. These will be fields 4-10.

```{r}
players_per_round <- lapply(first_lines_of_records_separate_fields,"[",4:10)
```

Now, use the word function from stringr to get the last "word" of each value in player_per_round.

We combine this with the trimws function so we will always extract either a number or a letter, never an empty string.

```{r}
players_per_round <- lapply(players_per_round,function(x)word(trimws(x),-1))
```

Finally, we extract only numbers, not letters.

Use the "+" to extract two-digit numbers as one string.

```{r}
players_per_round <- lapply(players_per_round,function(x)unlist(str_extract_all(x,'\\d+')))
```

Let's paste this together with the original strings to check our work.

```{r}
original_strings_for_players_per_round <- lapply(first_lines_of_records_separate_fields,"[",4:10)
original_strings_for_players_per_round <- unlist(lapply(original_strings_for_players_per_round,function(x)paste0(x,collapse=",")))
players_per_round_as_vector <- unlist(lapply(players_per_round,function(x)paste0(x,collapse=",")))
data.frame(Original.string = original_strings_for_players_per_round,New.string = players_per_round_as_vector)
```

Looks good!

## Putting it all together

Now, we need to use players_per_round to get the average pre-rating for the other players.

Then, combine this information with all the other information we need.

```{r}
ratings <- as.numeric(ratings)

mean_rating_other_players <- c()

for(i in 1:length(players_per_round))
{
players_per_round[[i]] <- as.numeric(players_per_round[[i]])
mean_rating_other_players <- c(mean_rating_other_players,mean(ratings[players_per_round[[i]]]))
}

mean_rating_other_players <- round(mean_rating_other_players,digits=0)

data_frame_for_csv <- data.frame(Player.name = player_names_reformatted,
	State = player_states,
	Total.points = as.numeric(total_points),
	Pre.rating = ratings,
	Average.pre.rating.opponents = mean_rating_other_players,
	stringsAsFactors=FALSE)

#Increase display width a bit so we can see everything side-by-side.
options(width=100)
data_frame_for_csv
```

Finally, let's change column names, then print to a CSV file.

```{r}
colnames(data_frame_for_csv) <- c("Player’s Name","Player’s State","Total Number of Points","Player’s Pre-Rating","Average Pre-Chess Rating of Opponents")
write.table(data_frame_for_csv,
	file="tournamentinfo_processed.csv",
	row.names=FALSE,col.names=TRUE,quote=TRUE,sep=",")
```
